---
title: "Code for Full Statistical Analysis"
output: 
  pdf_document:
    keep_tex: false
date: "2024-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(kableExtra)
library(geepack)
```

```{r}
# Read the raw data 
vaccine <- read.csv("vaccine.csv") 

# States info that were collected in 2017-18 
state2017 <- c("Colorado", "Connecticut", "Minnesota", "Montana", "New Jersey", 
               "New York", "North Dakota", "Pennsylvania", "South Dakota", 
               "Utah", "Washington")

# State abbreviations mapping
state_abbreviations <- c(
  "Alabama"="AL", "Alaska"="AK", "Arizona"="AZ", "Arkansas"="AR", "California"="CA",
  "Colorado"="CO", "Connecticut"="CT", "Delaware"="DE", "Florida"="FL", "Georgia"="GA",
  "Hawaii"="HI", "Idaho"="ID", "Illinois"="IL", "Indiana"="IN", "Iowa"="IA",
  "Kansas"="KS", "Kentucky"="KY", "Louisiana"="LA", "Maine"="ME", "Maryland"="MD",
  "Massachusetts"="MA", "Michigan"="MI", "Minnesota"="MN", "Mississippi"="MS",
  "Missouri"="MO","Montana"="MT", "Nebraska"="NE", "Nevada"="NV", "New Hampshire"="NH",
  "New Jersey"="NJ","New Mexico"="NM", "New York"="NY", "North Carolina"="NC", 
  "North Dakota"="ND", "Ohio"="OH","Oklahoma"="OK", "Oregon"="OR", "Pennsylvania"="PA",
  "Rhode Island"="RI", "South Carolina"="SC", "South Dakota"="SD", "Tennessee"="TN",
  "Texas"="TX", "Utah"="UT", "Vermont"="VT", "Virginia"="VA", "Washington"="WA",
  "West Virginia"="WV", "Wisconsin"="WI", "Wyoming"="WY"
)
```


# Data Cleaning 

```{r}
vac <- vaccine %>% 
  mutate(type = ifelse(type == "", "Unknown", type), 
         mmr = ifelse(mmr == -1, NA, mmr), 
         overall = ifelse(overall == -1, NA, overall), 
         per_capita = statespending2016/schagepop2016, 
         year = case_when(state %in% state2017 ~ "2017-18", TRUE ~ "2018-19"), 
         state_abbr = state_abbreviations[state], 
         county = ifelse(county == "", NA, county), 
         city = ifelse(city == "", NA, city)) %>% 
  ungroup()
```


### Check School Types 

```{r}
vac %>% group_by(year, type) %>% summarise(n=n()) %>% ungroup() %>% 
  pivot_wider(names_from = "year", values_from = "n")
```

Regroup school type by assigning BOCES, kindergarten, nonpublic to "Other" category. 

```{r}
data <- vac %>% mutate(type = case_when(type %in% c("BOCES", "Kindergarten", "Nonpublic") ~ "Others", TRUE ~ type))
```


### Check missings in vaccination rate 

```{r}
# Summary table for missing vaccination rates by states 
rate_na_df <- data %>% group_by(state) %>% 
  summarise(state_abb = unique(state_abbr), mmr_na = sum(is.na(mmr)), overall_na=sum(is.na(overall)), 
            total_na = mmr_na+overall_na, N = n()) %>% 
  arrange(total_na)

# Produce a table for reporting the number of schools without rate reports
rate_na_df %>% select(-state_abb) %>% 
  kable(booktabs = TRUE, escape = FALSE, 
        col.names = c("State", "MMR", "Overall", "Total", "Total No.Schools")) %>% 
  add_header_above(c(" "=1, "Number of Missing in Vaccination Rate" = 3, " "=1)) %>% 
  kable_styling(latex_options = c("striped", "scale_down"))

no_overall <- rate_na_df %>% filter(overall_na==N) %>% pull(state_abb)
no_mmr <- rate_na_df %>% filter(mmr_na==N) %>% pull(state_abb)
```

```{r}
# Reshape data for plotting
total_na_long <- data %>% group_by(state_abbr) %>% 
  mutate(has_NArate = ifelse(is.na(mmr)&is.na(overall), 1, 0)) %>% 
  summarise(missingN = sum(has_NArate), N = n()) %>% 
  arrange(missingN) %>% 
  pivot_longer(cols = c(missingN, N), names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = c("missingN", "N"), labels = c("Total Missing", "Total Schools")))

# Creating the grouped bar plot
ggplot(total_na_long, aes(x = reorder(state_abbr, -Value), y = Value, fill = Variable)) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Total Missing" = "steelblue", "Total Schools" = "lightgray")) +
  labs(title = "Comparison of Total Missing Data and Total Schools by State",
       x = "State",
       y = "Count") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(size = 14), # X axis labels
    axis.text.y = element_text(size = 14), # Y axis labels
    axis.title.x = element_text(size = 16), # X axis title
    axis.title.y = element_text(size = 16), # Y axis title
    plot.title = element_text(size = 20, face = "bold"), # Plot title
    legend.title = element_text(size = 16), # Legend title
    legend.text = element_text(size = 14), # Legend items
    legend.position = "bottom"
  )

# ggsave("totalNA.png", width = 12, height = 9, dpi = 300)
```


```{r}
# MMR rate
mmrNA_long <- data %>% group_by(state_abbr) %>% 
  summarise(N = n(), mmr_na = sum(is.na(mmr))) %>%
  filter(mmr_na != 0) %>% 
  arrange(mmr_na) %>% 
  pivot_longer(cols = c(mmr_na, N), names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = c("mmr_na", "N"), 
                           labels = c("MMR Missing", "Total Schools")))

mmrNA_long %>% 
  ggplot(aes(x = reorder(state_abbr, -Value), y = Value, fill = Variable)) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Total Schools" = "lightgray", "MMR Missing" = "steelblue")) +
  labs(title = "Missing MMR Vaccination Rates by State",
       x = "State",
       y = "Total Number of Missing Values") +
  theme_minimal() +theme(axis.text.x = element_text(size = 14), # X axis labels
        axis.text.y = element_text(size = 14), # Y axis labels
        axis.title.x = element_text(size = 16), # X axis title
        axis.title.y = element_text(size = 16), # Y axis title
        plot.title = element_text(size = 20, face = "bold"), 
        legend.title = element_text(size = 16), # Legend title
        legend.text = element_text(size = 14), # Legend items
        legend.position = "bottom")
ggsave("plots/MMR_NA.png", width = 10, height = 8, dpi = 300)


# Overall rate
overallNA_long <- data %>% group_by(state_abbr) %>% 
  summarise(N = n(), overall_na = sum(is.na(overall))) %>%
  filter(overall_na != 0) %>% 
  arrange(overall_na) %>% 
  pivot_longer(cols = c(overall_na, N), names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = c("overall_na", "N"), 
                           labels = c("Missing in Overall", "Total Schools")))
overallNA_long %>% 
  ggplot(aes(x = reorder(state_abbr, -Value), y = Value, fill = Variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Total Schools" = "lightgray", 
                               "Missing in Overall" = "steelblue")) +
  labs(title = "Missing Overall Vaccination Rates by State",
       x = "State",
       y = "No.Missing in Overall Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 14), # X axis labels
        axis.text.y = element_text(size = 14), # Y axis labels
        axis.title.x = element_text(size = 16), # X axis title
        axis.title.y = element_text(size = 16), # Y axis title
        plot.title = element_text(size = 20, face = "bold"), 
        legend.title = element_text(size = 16), # Legend title
        legend.text = element_text(size = 14), # Legend items
        legend.position = "bottom") # Plot title
ggsave("plots/overall_NA.png", width = 10, height = 8, dpi = 300)
```



### Check Duplicates 

```{r}
obsID <- 1:nrow(data)
data <- data %>% mutate(oID = obsID) %>% relocate(oID, .before =state)
duplicate_df <- data %>% 
  group_by(state, county, city, name) %>% 
  reframe(oID = oID, type = type, year = year, n= n(), enroll = enroll, mmr = mmr, overall = overall) %>% 
  filter(n > 1) %>% 
  group_by(state, county, city, name) %>% 
  mutate(record = row_number()) %>% ungroup()

duplicate_df %>% 
  filter(state == "California") %>% 
  select(oID, record, name, county, city, type, year, enroll, mmr, overall) %>% 
  kable(booktabs = TRUE, 
        col.names = c("School ID", "No.Record", "School Name", "County", "City","Type", "Year", 
                      "Enrollment","MMR Rate", "Overall Rate"), 
        caption = "Sample Duplicated School Observations in California") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  landscape()

# Calculate average of duplicated data 
avg_duplicates <- duplicate_df %>%
  group_by(state, county, city, name) %>% 
  mutate(mean_enroll = mean(enroll), mean_mmr = mean(mmr), mean_overall = mean(overall)) %>% 
  select(-c(mmr, overall, enroll, n)) %>% 
  rename(mmr = mean_mmr, overall=mean_overall, enroll = mean_enroll)

remove_sid <- avg_duplicates %>% filter(record != 1) %>% pull(oID)

data <- data %>% filter(!(oID %in% remove_sid)) 
clean_data <- data %>% 
  # Add the total number of schools in each state 
  left_join(data %>% group_by(state) %>% summarise(n_school = n_distinct(name)), by = "state")
```

\newpage

Table 1 

```{r}
clean_data %>% 
  group_by(state) %>% 
  summarise(overall = n(), n_city = n_distinct(city), 
            n_county = n_distinct(county), 
            pop2016 = unique(schagepop2016), 
            expenditure = unique(statespending2016), 
            capita = unique(per_capita)) %>% 
  kable(booktabs = TRUE, 
        col.names = c("State", "N (Schools)", "N (City)", "N (County)", 
                      "School-Aged\nPopulation (2016)", 
                      "Expenditure (2016)", "Per Capita")) %>% 
  kable_styling(latex_options = c("striped", "scale_down")) 
```


```{r}
clean_data %>% 
  group_by(state) %>% 
  summarise(n_city = n_distinct(city), n_county = n_distinct(county), 
            n_cityNA = sum(is.na(city)), n_countyNA = sum(is.na(county))) %>% 
  mutate(has_both = ifelse(n_cityNA == 0 & n_countyNA == 0, 1, 0), 
         only_city = ifelse(n_cityNA == 0 & n_countyNA !=0, 1, 0), 
         only_county = ifelse(n_countyNA == 0 & n_cityNA !=0, 1, 0), 
         none = ifelse(n_city == 1 & n_county == 1, 1, 0), 
         other = ifelse(n_city > 1 & n_county > 1, 1, 0)) %>% 
  summarise(N_both = sum(has_both), Ncityonly = sum(only_city), 
            Ncountyonly = sum(only_county), N_none = sum(none), 
            Nothers = sum(other))
```


## Filter dataframes for MMR and Overall Vaccination Rate

```{r}
state_rate_df <- clean_data %>% 
  group_by(state) %>% 
  summarise(n_school = unique(n_school), 
            mmr_na = sum(is.na(mmr)),
            n_mmr = sum(!is.na(mmr)),
            overall_na = sum(is.na(overall)), 
            n_overall = sum(!is.na(overall)))

overall_states <- state_rate_df %>% filter(overall_na < n_overall) %>% pull(state)
mmr_states <- state_rate_df %>% filter(mmr_na < n_mmr) %>% pull(state)

overall_df <- clean_data %>% filter(state %in% overall_states)
mmr_df <- clean_data %>% filter(state %in% mmr_states)
```

## Stratified Sampling 

### MMR 

```{r}
n_states <- n_distinct(mmr_df$state)
mmr_dat <- mmr_df %>% 
  filter(!is.na(mmr)) %>%  # remove all the schools without MMR rate 
  mutate(year = case_when(year == "2017-18" ~ 0, TRUE ~ 1)) %>% # Convert year to an indicator column
  left_join(mmr_df %>% group_by(state) %>% slice(1) %>%
              cbind(stateID = 1:n_states) %>%  # Create an ID column for each state 
              select(state, stateID), by = "state") %>% 
  relocate(stateID, .before = oID) 

# Stratified Sampling 
# Step 1: Calculate weights for each state based on the number of schools
state_weights <- mmr_dat %>%
  count(state) %>%
  mutate(weight = n / sum(n))

# Step 2: Sample schools within each state, proportional to state weights
# Set a total sample size of N schools
N <- 3000  # Total desired sample size

# Calculate the number of schools to sample from each state, based on weights
state_weights <- state_weights %>%
  mutate(sample_size = round(weight * N))

set.seed(123)

# mmr_samples <- mmr_dat %>%
#   group_by(state) %>%
#   sample_n(N, replace = FALSE) %>%
#   mutate(schoolID = row_number()) %>%
#   relocate(schoolID, .after = stateID) %>%
#   ungroup() %>%
#   arrange(stateID)


mmr_samples <- NULL
for (i in 1:nrow(state_weights)) {
  state <- state_weights[i, ]$state
  size <- state_weights$sample_size[i]
  sample <- mmr_dat %>% filter(state == state) %>%
    sample_n(size, replace = FALSE)
  mmr_samples <- rbind(mmr_samples, sample)
}

mmr_samples <- mmr_samples %>% group_by(state) %>%
  mutate(schoolID = row_number()) %>%
  relocate(schoolID, .after = stateID) %>%
  ungroup() %>%
  arrange(stateID)
```


## Cleaning for Overall Vaccination 

```{r}
n_states <- n_distinct(overall_df$state)
overall_dat <- overall_df %>% 
  filter(!is.na(mmr)) %>%  # remove all the schools without MMR rate 
  mutate(year = case_when(year == "2017-18" ~ 0, TRUE ~ 1)) %>% # Convert year to an indicator column
  left_join(overall_df %>% group_by(state) %>% slice(1) %>%
              cbind(stateID = 1:n_states) %>%  # Create an ID column for each state 
              select(state, stateID), by = "state") %>% 
  relocate(stateID, .before = oID) 

# Stratified Sampling 
# Step 1: Calculate weights for each state based on the number of schools
state_weights <- overall_dat %>%
  count(state) %>%
  mutate(weight = n / sum(n))

# Step 2: Sample schools within each state, proportional to state weights
# Set a total sample size of N schools
N <- 3000  # Total desired sample size

# Calculate the number of schools to sample from each state, based on weights
state_weights <- state_weights %>%
  mutate(sample_size = round(weight * N))

set.seed(10049)
overall_samples <- NULL
for (i in 1:nrow(state_weights)) {
  state <- state_weights[i, ]$state
  size <- state_weights$sample_size[i]
  sample <- overall_dat %>% filter(state == state) %>% 
    sample_n(size, replace = FALSE)
  overall_samples <- rbind(overall_samples, sample)
}

overall_samples <- overall_samples %>% group_by(state) %>% 
  mutate(schoolID = row_number()) %>% 
  relocate(schoolID, .after = stateID) %>% 
  ungroup() %>% 
  arrange(stateID)
```



# EDA

\newpage 
State School number 
```{r}
clean_data %>% group_by(state) %>% select(state, n_school) %>% slice(1) %>% arrange(n_school) %>% 
  kable(booktabs = TRUE, col.names = c("State", "Schools (N)"), 
        caption = "The number of schools within each state.") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) 
```


## Data Visualization

```{r}
mmr_df %>% group_by(state) %>% 
  mutate(mean_rate = mean(mmr)) %>% 
  ggplot(aes(x = per_capita, y = mean_rate))+
  geom_point(shape = 1)+
  scale_colour_hue(l=50)+
  geom_smooth(method = lm)+
  labs(x = "Per Capita (2016)", y = "State Level Average MMR Rate")

overall_df %>% group_by(state) %>% 
  mutate(mean_rate = mean(overall)) %>% 
  ggplot(aes(x = per_capita, y = mean_rate))+
  geom_point(shape = 1)+
  scale_colour_hue(l=50)+
  geom_smooth(method = lm)+
  labs(x="Per Capita (2016)", y = "State Level Overall Rate")


```



### Vaccination Rates by States 

```{r}
rate_long <- clean_data %>% 
  select(state_abbr, mmr, overall) %>%
  pivot_longer(cols = c(mmr, overall), 
               names_to = "vaccine_type", values_to = "rate") %>% 
  mutate(Vaccine = case_when(vaccine_type == "mmr" ~ "MMR", TRUE ~ "Overall"))

rate_long %>% 
  filter(state_abbr %in% no_mmr) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "Overall Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(state_abbr %in% no_overall) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "MMR Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(!(state_abbr %in% c(no_mmr, no_overall))) %>% 
  mutate(Vaccine = factor(Vaccine)) %>% 
  ggplot(aes(x = state_abbr, y = rate)) +
  geom_boxplot(aes(fill = Vaccine)) +
  labs(x = "State", y = "Rate", 
       title = "Comparison of Vaccination Rate Distribution By Type") +
  scale_fill_brewer(palette = 1)+
  theme_minimal()+
  theme(axis.text.x = element_text(size = 14), # X axis labels
        axis.text.y = element_text(size = 14), # Y axis labels
        axis.title.x = element_text(size = 16), # X axis title
        axis.title.y = element_text(size = 16), # Y axis title
        plot.title = element_text(size = 20, face = "bold"), 
        legend.title = element_text(size = 16), # Legend title
        legend.text = element_text(size = 14), # Legend items
        legend.position = "bottom") # Plot title

# ggsave("plots/vac_comp.png",  width = 10, height = 8, dpi = 300)
```


### State School Aged Population and Per Capita

```{r}
clean_data %>% 
  ggplot(aes(x = reorder(state_abbr, -schagepop2016), y = schagepop2016)) +
  geom_point(size = 3, color = "red") + 
  geom_segment(aes(x = state_abbr, xend = state_abbr, y = 0, yend = schagepop2016)) +
  labs(x = "State",y= "School-Aged Population in 2016")+
  theme_minimal() +
  theme(axis.text.x = element_text(size = 14), # X axis labels
        axis.text.y = element_text(size = 14)) # Plot title
# ggsave("plots/StateschAged.png",  width = 10, height = 8, dpi = 300)

clean_data %>% 
  ggplot(aes(x = reorder(state_abbr, -per_capita), y = per_capita)) +
  geom_point(size = 3, color = "red") + 
  geom_segment(aes(x = state_abbr, xend = state_abbr, y = 0, yend = per_capita)) +
  labs(x = "State", y="Per Capita")+
  theme_minimal()+
  theme(axis.text.x = element_text(size = 14), # X axis labels
        axis.text.y = element_text(size = 14)) # Plot title
# ggsave("plots/StateCapita.png",  width = 10, height = 8, dpi = 300)
```


### Vaccination rate vs. School, County, City, Capita, and Spending 

```{r}
state_summary <- clean_data %>% group_by(state) %>% 
  summarise(Total_schools = n(), 
            n_county = ifelse(sum(is.na(county)) == Total_schools, 0, n_distinct(county)), 
            n_county_na = sum(is.na(county)),
            n_city = ifelse(sum(is.na(city)) == Total_schools, 0, n_distinct(city)), 
            n_city_na = sum(is.na(city)), 
            avg_mmr = mean(mmr, na.rm = TRUE), 
            avg_overall = mean(overall, na.rm = TRUE), 
            avg_enroll = mean(enroll)) %>% 
  arrange(avg_enroll) %>% 
  left_join(clean_data %>% group_by(state) %>% 
          select(state, year, per_capita, schagepop2016, statespending2016), by="state")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_mmr, color =year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "State Spending in 2016", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "State Spending 2016", 
       y = "Average Overall Vaccination Rate")
```

### Florida 

```{r}
clean_data %>% filter(state == "Florida") %>% select(overall) %>% summary()
```


## EDA for MMR Data 

```{r}
mmr_df %>% 
  group_by(state) %>% 
  summarise(n_school = n_distinct(name),
            min_mmr = min(mmr, na.rm=TRUE),
            mean_mmr = mean(mmr, na.rm=TRUE), 
            median_mmr = median(mmr, na.rm=TRUE), 
            max_mmr = max(mmr, na.rm=TRUE), 
            mmr_na = sum(is.na(mmr)), 
            year = unique(year)) %>% 
  arrange(year) %>% 
  kable(booktabs = TRUE, caption = "The summary of MMR vaccination rates", 
        col.names = c("State", "No.School", "Min", "Mean", "Median", "Max", "No.Missing", "School Year")) %>% 
  kable_styling(latex_options = c("striped", "scale_down")) 
```



## EDA for Overall Data 

```{r}
overall_df %>% 
  group_by(state) %>% 
  summarise(n_school = n_distinct(name),
            min_overall = min(overall, na.rm=TRUE),
            mean_overall = mean(overall, na.rm=TRUE), 
            median_overall = median(overall, na.rm=TRUE), 
            max_overall = max(overall, na.rm=TRUE), 
            overall_na = sum(is.na(overall)), 
            year = unique(year)) %>% 
  arrange(year) %>% 
  kable(booktabs = TRUE, caption = "The summary of Overall vaccination rates", 
        col.names = c("State", "No.School", "Min", "Mean", "Median", "Max", "No.Missing", "School Year")) %>% 
  kable_styling(latex_options = c("striped", "scale_down")) 
```



# Statistical Analysis 

The data has many schools within each state, meaning that each state has repeated measurements by schools. Therefore, the vaccination rate between schools within the same state are not independent of each other, violating the assumption of independence of standard regression models. Therefore, we will model the covariance structures with Generalized Least Squares. 

## MMR 

Let $Y_{ij}$ denote the MMR vaccination rate of the $i^{th}$ state and $j^{th}$ school. 

$$
E(Y_{ij}) = \beta_0 + \beta_1 \text{Capita}_i + \beta_2\text{NSchool}_i+\beta_3 \text{Year}_i
$$
where Capita$_i$ was calculated by dividing the state spending by school aged population, `NSchool` represents the number of schools in each state, and `Year` denotes the school-year that the observation is recorded (1 = 2018-19, 0 = 2017-18). 


### Generalized Least Squares 

Failed 

```{r, eval=FALSE}
library(nlme)
# Unstructured
mmr_unstr <- gls(mmr ~ per_capita+n_school+year, 
                 corr=corSymm(form = ~ schoolID | stateID), 
                 weights = varIdent(form = ~ 1|schoolID), 
                 data = mmr_samples)

getVarCov(mmr_unstr)

# Compound Symmetry (heterogeneous)
mmr_compsym <- gls(mmr ~ per_capita+n_school+year, 
                   corr=corCompSymm(form = ~ schoolID | stateID), 
                 weights = varIdent(form = ~ 1|schoolID), 
                 data = mmr_samples)
getVarCov(mmr_compsym)

# AR1 (heterogeneous)
mmr_ar1 <- gls(mmr ~ per_capita+n_school+year, 
                   corr=corAR1(form = ~ schoolID | stateID), 
                 weights = varIdent(form = ~ 1|schoolID), 
                 data = mmr_samples)
getVarCov(mmr_ar1)

# Exponential 
mmr_expo <- gls(mmr ~ per_capita+n_school+year, 
                   corr=corExp(form = ~ schoolID | stateID), 
                 weights = varIdent(form = ~ 1|schoolID), 
                 data = mmr_samples)
getVarCov(mmr_expo)

#### Model comparison
outAIC <- function(x){
  sumx <- summary(x)
  out <- c(sumx$AIC, sumx$BIC)
  names(out) <- c("AIC", "BIC")
  out
}

allaic <- rbind(
  Unstructured = outAIC(mmr_unstr), 
  CompoundSymmetry = outAIC(mmr_compsym), 
  AR1 = outAIC(mmr_ar1), 
  Exponential = outAIC(mmr_expo)
)

allaic
```


### Unadjusted GEE 

$$
\begin{aligned}
\text{MMR}_{ij} &= \beta_0 +\beta_1 \cdot \text{PCS}_{ij} + \beta_2 \cdot \text{NSch}_{ij}+\beta_3 \cdot \text{Type}_{ij} + \epsilon_{ij}\\
\text{OVR}_{ij} &= \beta_0 +\beta_1 \cdot \text{PCS}_{ij} + \beta_2 \cdot \text{NSch}_{ij}+\beta_3 \cdot \text{Type}_{ij} + \epsilon_{ij}
\end{aligned}
$$

$$
\begin{bmatrix}
1 & 0 & 0 &\cdots &0\\
0 & 1 & 0 &\cdots &0\\
0 & 0 & 1 &\cdots &0\\
\vdots & \vdots &\vdots &\ddots & \vdots\\
0 & 0 & 0 &\cdots &1
\end{bmatrix}
$$

```{r}
# GEE with Exchangeable Structure
mmr_ex <- geeglm(mmr ~ per_capita, family = gaussian, 
                  data = mmr_samples, id = stateID, 
                  corstr = "exchangeable")

# GEE with AR1 Structure
mmr_ar1 <- geeglm(mmr ~ per_capita, family = gaussian, 
                  data = mmr_samples, id = stateID, 
                  corstr = "ar1")

# GEE with Independent Structure
mmr_ind <- geeglm(mmr ~ per_capita, family = gaussian, 
                  data = mmr_samples, id = stateID, 
                  corstr = "independence")

# Model comparison 
anova(mmr_ex, mmr_ar1)
anova(mmr_ex, mmr_ind)
anova(mmr_ind, mmr_ar1)

# QIC Measures of Fit 
MuMIn::model.sel(mmr_ex, mmr_ar1, mmr_ind, rank="QIC")
```

1. ANOVA suggests that the three unadjusted models are all identical. 

2. Exchangeable Structure has the lowest QIC. 


**Statistical Metrics:**

* qLike: Quasi-likelihood of the model, a measure of the model fit to the data. Higher values (closer to zero, since these are likely negative) generally indicate a better fit.  
* QIC: Quasi Information Criterion, a measure used for model selection. Lower QIC values indicate a better model fit considering both the goodness of fit and the complexity of the model.  
* Delta: This column shows the difference in QIC from the best model to the current model. A delta of 0 indicates the best model.  
* Weight: These are the Akaike weights based on the QIC values, providing a measure of each model's relative likelihood of being the best model among the set of models evaluated. A higher weight suggests a higher likelihood that this model is the best among those tested. 

### Adjusted GEE 

```{r}
# Adjusted GEE with exchangeable 
mmr_adj_ex <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "exchangeable")

# Adjusted GEE with AR1 
mmr_adj_ar1 <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "ar1")

mmr_adj_ind <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "independence")

anova(mmr_adj_ex, mmr_adj_ar1)
anova(mmr_adj_ex, mmr_adj_ind)
anova(mmr_adj_ar1, mmr_adj_ind)

# QIC Measures of Fit 
MuMIn::model.sel(mmr_adj_ex, mmr_adj_ar1, mmr_adj_ind, rank="QIC")
```


```{r}
data.frame(
  struc = c("Exchangeable", "AR1", "Independence"), 
  QIC = c(3007.5, 3015.7, 3015.8, 
          3027.8, 3046.3, 3046.6), 
  delta = c(0.00, 8.25, 8.26, 
            0.00, 18.54, 18.86), 
  weight = c(0.969, 0.016, 0.016, 
             1, 0, 0)) %>% 
  kable(booktabs = TRUE, 
        col.names = c("Structure", "QIC", "Delta", "Weight"),
        caption = "Model Selection Result for MMR Rate") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  pack_rows("Unadjusted", 1, 3) %>% 
  pack_rows("Adjusted", 4, 6)
```

```{r}
anova(mmr_ex, mmr_adj_ex)
```

```{r}
summary(mmr_adj_ex)
```


## Overall Unadjusted 

```{r}
# Exchangeable
unadj_overall_ex <- geeglm(overall ~ per_capita, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "exchangeable")

# AR1
unadj_overall_ind <- geeglm(overall ~ per_capita, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "independence")

# Fit the GEE model
unadj_overall_ar1 <- geeglm(overall ~ per_capita, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "ar1")

anova(unadj_overall_ex, unadj_overall_ar1)
anova(unadj_overall_ex, unadj_overall_ind)
anova(unadj_overall_ar1, unadj_overall_ind)
```
```{r}
# QIC Measures of Fit 
# Take long time to run 
# MuMIn::model.sel(unadj_overall_ex, unadj_overall_ind, unadj_overall_ar1, rank="QIC")
```

```{r}
data.frame(
  struc = c("Exchangeable", "Independence", "AR1"), 
  QIC = c(2934.4, 3012.3, 3012.3), 
  delta = c(0.00, 77.92, 77.94), 
  weight = c(1, 0, 0)) %>% 
  kable(booktabs = TRUE, 
        col.names = c("Structure", "QIC", "Delta", "Weight"),
        caption = "Unadjusted Model Selection Result for Overall Rate") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) 
```

```{r}
# Exchangeable
adj_overall_ex <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "exchangeable")

# Independence
adj_overall_ind <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "independence")

# AR1
adj_overall_ar1 <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "ar1")

anova(adj_overall_ex, adj_overall_ind)
anova(adj_overall_ex, adj_overall_ar1)
anova(adj_overall_ar1, adj_overall_ind)
```

```{r}
# QIC Measures of Fit 
MuMIn::model.sel(adj_overall_ex, adj_overall_ind, adj_overall_ar1, rank="QIC")
```


```{r}
data.frame(
  struc = c("AR1", "Independence", "Exchangeable"), 
  QIC = c(3062.3, 3062.4, 3065.7), 
  delta = c(0.00, 0.05, 3.34), 
  weight = c(0.462, 0.451, 0.087)) %>% 
  kable(booktabs = TRUE, 
        col.names = c("Structure", "QIC", "Delta", "Weight"),
        caption = "Adjusted Model Selection Result for Overall Rate") %>% 
  kable_styling(latex_options = c("striped", "scale_down"))
```


```{r}
anova(unadj_overall_ex, adj_overall_ar1)
```

```{r}
summary(adj_overall_ar1)$coefficients
```


```{r}
data.frame(
  ` ` = c("Intercept", "Per Capita", "Number of Schools", 
          "Private", "Public", "Unknown"), 
  Estimate = c(79.06, 0.000813, 0.00128, -4.11, -2.04, -2.96), 
  Std.err = c(4.364, 0.000496, 0.000106, 1.324, 1.260, 3.955),
  Wald = c(328.197, 2.688, 145.729, 9.642, 2.611, 0.559), 
  `P-value` = c("< 0.05", "0.101", "<0.05", "<0.05", "0.106", "0.454"))%>% 
  kable(booktabs = TRUE, 
        caption = "Best Model Estimation Result for MMR Vaccination Rate") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  pack_rows("School Type", 4, 6) 
```





