---
title: "Jinyu's Draft"
output: 
  pdf_document:
    keep_tex: false
date: "2024-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(kableExtra)
```

```{r}
vaccine <- read.csv("vaccine.csv")
```

```{r}
summary(vaccine)
```

# Data Cleaning 

```{r}
table(vaccine$type)
```

```{r}
state2017 <- c("Colorado", "Connecticut", "Minnesota", "Montana", "New Jersey", 
               "New York", "North Dakota", "Pennsylvania", "South Dakota", 
               "Utah", "Washington")

# State abbreviations mapping
state_abbreviations <- c(
  "Alabama"="AL", "Alaska"="AK", "Arizona"="AZ", "Arkansas"="AR", "California"="CA",
  "Colorado"="CO", "Connecticut"="CT", "Delaware"="DE", "Florida"="FL", "Georgia"="GA",
  "Hawaii"="HI", "Idaho"="ID", "Illinois"="IL", "Indiana"="IN", "Iowa"="IA",
  "Kansas"="KS", "Kentucky"="KY", "Louisiana"="LA", "Maine"="ME", "Maryland"="MD",
  "Massachusetts"="MA", "Michigan"="MI", "Minnesota"="MN", "Mississippi"="MS", "Missouri"="MO",
  "Montana"="MT", "Nebraska"="NE", "Nevada"="NV", "New Hampshire"="NH", "New Jersey"="NJ",
  "New Mexico"="NM", "New York"="NY", "North Carolina"="NC", "North Dakota"="ND", "Ohio"="OH",
  "Oklahoma"="OK", "Oregon"="OR", "Pennsylvania"="PA", "Rhode Island"="RI", "South Carolina"="SC",
  "South Dakota"="SD", "Tennessee"="TN", "Texas"="TX", "Utah"="UT", "Vermont"="VT",
  "Virginia"="VA", "Washington"="WA", "West Virginia"="WV", "Wisconsin"="WI", "Wyoming"="WY"
)

vac <- vaccine %>% 
  mutate(type = ifelse(type == "", "Unknown", type), 
         mmr = ifelse(mmr == -1, NA, mmr), 
         overall = ifelse(overall == -1, NA, overall), 
         per_capita = statespending2016/schagepop2016, 
         year = case_when(state %in% state2017 ~ "2017-18", TRUE ~ "2018-19"), 
         state_abbr = state_abbreviations[state], 
         county = ifelse(county == "", NA, county), 
         city = ifelse(city == "", NA, city))
```

```{r}
vac %>% group_by(year, type) %>% summarise(n=n()) %>% ungroup() %>% 
  pivot_wider(names_from = "year", values_from = "n")
```

#### Inconsistency in school type. 

Regroup school type by assigning BOCES, kindergarten, nonpublic to "Other" category. 
```{r}
data <- vac %>% mutate(type = case_when(type %in% c("BOCES", "Kindergarten", "Nonpublic") ~ "Others", TRUE ~ type))
```


```{r}
rate_na_df <- data %>% group_by(state) %>% 
  summarise(state_abb = unique(state_abbr), mmr_na = sum(is.na(mmr)), overall_na=sum(is.na(overall)), 
            total_na = mmr_na+overall_na, N = n()) %>% 
  arrange(total_na)

rate_na_df %>% select(-state_abb) %>% 
  kable(booktabs = TRUE, escape = FALSE, 
        col.names = c("State", "MMR", "Overall", "Total", "Total No.Schools")) %>% 
  add_header_above(c(" "=1, "Number of Missing in Vaccination Rate" = 3, " "=1)) %>% 
  kable_styling(latex_options = c("striped", "scale_down"))

no_overall <- rate_na_df %>% filter(overall_na==N) %>% pull(state_abb)
no_mmr <- rate_na_df %>% filter(mmr_na==N) %>% pull(state_abb)
```


## Duplicates 

```{r}
schoolID <- 1:nrow(data)
data <- data %>% mutate(sID = schoolID) %>% relocate(sID, .before =state)
duplicate_df <- data %>% 
  group_by(state, county, city, name) %>% 
  reframe(sID = sID, type = type, year = year, n= n(), enroll = enroll, mmr = mmr, overall = overall) %>% 
  filter(n > 1) %>% 
  group_by(state, county, city, name) %>% 
  mutate(record = row_number()) %>% ungroup()

duplicate_df %>% 
  filter(state == "California") %>% 
  select(sID, record, name, county, city, type, year, enroll, mmr, overall) %>% 
  kable(booktabs = TRUE, 
        col.names = c("School ID", "No.Record", "School Name", "County", "City","Type", "Year", 
                      "Enrollment","MMR Rate", "Overall Rate"), 
        caption = "Sample Duplicated School Observations in California") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  landscape()

# Calculate average of duplicated data 
avg_duplicates <- duplicate_df %>%
  group_by(state, county, city, name) %>% 
  mutate(mean_enroll = mean(enroll), mean_mmr = mean(mmr), mean_overall = mean(overall)) %>% 
  select(-c(mmr, overall, enroll, n)) %>% 
  rename(mmr = mean_mmr, overall=mean_overall, enroll = mean_enroll)

remove_sid <- avg_duplicates %>% filter(record != 1) %>% pull(sID)

clean_data <- data %>% filter(!(sID %in% remove_sid))
clean_data$sID <- 1:nrow(clean_data)

```



# Data Visualization

## Vaccination Rates by States 

```{r}
rate_long <- clean_data %>% 
  select(state_abbr, mmr, overall) %>%
  pivot_longer(cols = c(mmr, overall), names_to = "vaccine_type", values_to = "rate")

rate_long %>% 
  filter(state_abbr %in% no_mmr) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "Overall Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(state_abbr %in% no_overall) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "MMR Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(!(state_abbr %in% c(no_mmr, no_overall))) %>% 
  ggplot(aes(y= rate, x = state_abbr, color = vaccine_type)) +
  geom_boxplot() +
  labs(x = "State", y = "rate_long") +
  scale_color_manual(values = c("overall" = "blue", "mmr" = "red")) + 
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))
```

# EDA 

```{r}
state_summary <- clean_data %>% group_by(state) %>% 
  summarise(Total_schools = n(), 
            n_county = ifelse(sum(is.na(county)) == Total_schools, 0, n_distinct(county)), 
            n_county_na = sum(is.na(county)),
            n_city = ifelse(sum(is.na(city)) == Total_schools, 0, n_distinct(city)), 
            n_city_na = sum(is.na(city)), 
            avg_mmr = mean(mmr, na.rm = TRUE), 
            avg_overall = mean(overall, na.rm = TRUE), 
            avg_enroll = mean(enroll)) %>% 
  arrange(avg_enroll) %>% 
  left_join(clean_data %>% group_by(state) %>% 
          select(state, year, per_capita, schagepop2016, statespending2016), by="state")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_mmr, color =year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "State Spending in 2016", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "State Spending 2016", 
       y = "Average Overall Vaccination Rate")
```

```{r}
final_data <- clean_data %>% 
  left_join(clean_data %>% group_by(state) %>% 
              summarise(n_school = n_distinct(name)))
write.csv(final_data, "vaccine_cleaned.csv", row.names=FALSE)
```

```{r}
clean_data %>% filter(state == "Florida") %>% select(overall) %>% summary()
```


# Statistical Analysis 

The data has many schools within each state, meaning that each state has repeated measurements by schools. Therefore, the vaccination rate between schools within the same state are not independent of each other, violating the assumption of independence of standard regression models. 

To account for correlation within each state, we will use GEEs which allow for the correlation of observations within clusters by specifying a working correlation structure. GEEs provide marginal effects (population-averaged), which can be more interpretative for policy analysis at the state level. 


```{r}
clean_data %>% group_by(state) %>% 
  summarise(avg_mmr = mean(mmr, na.rm = TRUE), 
            avg_overall = mean(overall, na.rm=TRUE))
prep_data <- clean_data %>%
  mutate(sch_year = case_when(year == "2017-18" ~ 1, 0),
         I_city = )
```
























