---
title: "Jinyu's Draft"
output: 
  pdf_document:
    keep_tex: false
date: "2024-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(kableExtra)
```

```{r}
# Read the raw data 
vaccine <- read.csv("vaccine.csv") 

# States info that were collected in 2017-18 
state2017 <- c("Colorado", "Connecticut", "Minnesota", "Montana", "New Jersey", 
               "New York", "North Dakota", "Pennsylvania", "South Dakota", 
               "Utah", "Washington")

# State abbreviations mapping
state_abbreviations <- c(
  "Alabama"="AL", "Alaska"="AK", "Arizona"="AZ", "Arkansas"="AR", "California"="CA",
  "Colorado"="CO", "Connecticut"="CT", "Delaware"="DE", "Florida"="FL", "Georgia"="GA",
  "Hawaii"="HI", "Idaho"="ID", "Illinois"="IL", "Indiana"="IN", "Iowa"="IA",
  "Kansas"="KS", "Kentucky"="KY", "Louisiana"="LA", "Maine"="ME", "Maryland"="MD",
  "Massachusetts"="MA", "Michigan"="MI", "Minnesota"="MN", "Mississippi"="MS", "Missouri"="MO",
  "Montana"="MT", "Nebraska"="NE", "Nevada"="NV", "New Hampshire"="NH", "New Jersey"="NJ",
  "New Mexico"="NM", "New York"="NY", "North Carolina"="NC", "North Dakota"="ND", "Ohio"="OH",
  "Oklahoma"="OK", "Oregon"="OR", "Pennsylvania"="PA", "Rhode Island"="RI", "South Carolina"="SC",
  "South Dakota"="SD", "Tennessee"="TN", "Texas"="TX", "Utah"="UT", "Vermont"="VT",
  "Virginia"="VA", "Washington"="WA", "West Virginia"="WV", "Wisconsin"="WI", "Wyoming"="WY"
)
```


# Data Cleaning 

```{r}
vac <- vaccine %>% 
  mutate(type = ifelse(type == "", "Unknown", type), 
         mmr = ifelse(mmr == -1, NA, mmr), 
         overall = ifelse(overall == -1, NA, overall), 
         per_capita = statespending2016/schagepop2016, 
         year = case_when(state %in% state2017 ~ "2017-18", TRUE ~ "2018-19"), 
         state_abbr = state_abbreviations[state], 
         county = ifelse(county == "", NA, county), 
         city = ifelse(city == "", NA, city)) %>% 
  ungroup()
```


### Check School Types 

```{r}
vac %>% group_by(year, type) %>% summarise(n=n()) %>% ungroup() %>% 
  pivot_wider(names_from = "year", values_from = "n")
```

Regroup school type by assigning BOCES, kindergarten, nonpublic to "Other" category. 

```{r}
data <- vac %>% mutate(type = case_when(type %in% c("BOCES", "Kindergarten", "Nonpublic") ~ "Others", TRUE ~ type))
```


### Check missings in vaccination rate 

```{r}
rate_na_df <- data %>% group_by(state) %>% 
  summarise(state_abb = unique(state_abbr), mmr_na = sum(is.na(mmr)), overall_na=sum(is.na(overall)), 
            total_na = mmr_na+overall_na, N = n()) %>% 
  arrange(total_na)

rate_na_df %>% select(-state_abb) %>% 
  kable(booktabs = TRUE, escape = FALSE, 
        col.names = c("State", "MMR", "Overall", "Total", "Total No.Schools")) %>% 
  add_header_above(c(" "=1, "Number of Missing in Vaccination Rate" = 3, " "=1)) %>% 
  kable_styling(latex_options = c("striped", "scale_down"))

no_overall <- rate_na_df %>% filter(overall_na==N) %>% pull(state_abb)
no_mmr <- rate_na_df %>% filter(mmr_na==N) %>% pull(state_abb)
```


### Check Duplicates 

```{r}
obsID <- 1:nrow(data)
data <- data %>% mutate(oID = obsID) %>% relocate(oID, .before =state)
duplicate_df <- data %>% 
  group_by(state, county, city, name) %>% 
  reframe(oID = oID, type = type, year = year, n= n(), enroll = enroll, mmr = mmr, overall = overall) %>% 
  filter(n > 1) %>% 
  group_by(state, county, city, name) %>% 
  mutate(record = row_number()) %>% ungroup()

duplicate_df %>% 
  filter(state == "California") %>% 
  select(oID, record, name, county, city, type, year, enroll, mmr, overall) %>% 
  kable(booktabs = TRUE, 
        col.names = c("School ID", "No.Record", "School Name", "County", "City","Type", "Year", 
                      "Enrollment","MMR Rate", "Overall Rate"), 
        caption = "Sample Duplicated School Observations in California") %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  landscape()

# Calculate average of duplicated data 
avg_duplicates <- duplicate_df %>%
  group_by(state, county, city, name) %>% 
  mutate(mean_enroll = mean(enroll), mean_mmr = mean(mmr), mean_overall = mean(overall)) %>% 
  select(-c(mmr, overall, enroll, n)) %>% 
  rename(mmr = mean_mmr, overall=mean_overall, enroll = mean_enroll)

remove_sid <- avg_duplicates %>% filter(record != 1) %>% pull(oID)

data <- data %>% filter(!(oID %in% remove_sid)) 
clean_data <- data %>% 
  # Add the total number of schools in each state 
  left_join(data %>% group_by(state) %>% summarise(n_school = n_distinct(name)), by = "state")
```


## Filter dataframes for MMR and Overall Vaccination Rate

```{r}
state_rate_df <- clean_data %>% 
  group_by(state) %>% 
  summarise(n_school = unique(n_school), 
            mmr_na = sum(is.na(mmr)),
            n_mmr = sum(!is.na(mmr)),
            overall_na = sum(is.na(overall)), 
            n_overall = sum(!is.na(overall)))

overall_states <- state_rate_df %>% filter(overall_na < n_overall) %>% pull(state)
mmr_states <- state_rate_df %>% filter(mmr_na < n_mmr) %>% pull(state)

overall_df <- clean_data %>% filter(state %in% overall_states)
mmr_df <- clean_data %>% filter(state %in% mmr_states)
```

## MMR data 

```{r}
n_states <- n_distinct(mmr_df$state)
mmr_dat <- mmr_df %>% 
  filter(!is.na(mmr)) %>%  # remove all the schools without MMR rate 
  mutate(year = case_when(year == "2017-18" ~ 0, TRUE ~ 1)) %>% # Convert year to an indicator column
  left_join(mmr_df %>% group_by(state) %>% slice(1) %>%
              cbind(stateID = 1:n_states) %>%  # Create an ID column for each state 
              select(state, stateID), by = "state") %>% 
  relocate(stateID, .before = oID) 

# Stratified Sampling 
# Step 1: Calculate weights for each state based on the number of schools
state_weights <- mmr_dat %>%
  count(state) %>%
  mutate(weight = n / sum(n))

# Step 2: Sample schools within each state, proportional to state weights
# Set a total sample size of N schools
N <- 5000  # Total desired sample size

# Calculate the number of schools to sample from each state, based on weights
state_weights <- state_weights %>%
  mutate(sample_size = round(weight * N))

set.seed(123)

# mmr_samples <- mmr_dat %>%
#   group_by(state) %>%
#   sample_n(N, replace = FALSE) %>%
#   mutate(schoolID = row_number()) %>%
#   relocate(schoolID, .after = stateID) %>%
#   ungroup() %>%
#   arrange(stateID)


mmr_samples <- NULL
for (i in 1:nrow(state_weights)) {
  state <- state_weights[i, ]$state
  size <- state_weights$sample_size[i]
  sample <- mmr_dat %>% filter(state == state) %>%
    sample_n(size, replace = FALSE)
  mmr_samples <- rbind(mmr_samples, sample)
}

mmr_samples <- mmr_samples %>% group_by(state) %>%
  mutate(schoolID = row_number()) %>%
  relocate(schoolID, .after = stateID) %>%
  ungroup() %>%
  arrange(stateID)
```


## Cleaning for Overall Vaccination 

```{r}
n_states <- n_distinct(overall_df$state)
overall_dat <- overall_df %>% 
  filter(!is.na(mmr)) %>%  # remove all the schools without MMR rate 
  mutate(year = case_when(year == "2017-18" ~ 0, TRUE ~ 1)) %>% # Convert year to an indicator column
  left_join(overall_df %>% group_by(state) %>% slice(1) %>%
              cbind(stateID = 1:n_states) %>%  # Create an ID column for each state 
              select(state, stateID), by = "state") %>% 
  relocate(stateID, .before = oID) 

# Stratified Sampling 
# Step 1: Calculate weights for each state based on the number of schools
state_weights <- overall_dat %>%
  count(state) %>%
  mutate(weight = n / sum(n))

# Step 2: Sample schools within each state, proportional to state weights
# Set a total sample size of N schools
N <- 1000  # Total desired sample size

# Calculate the number of schools to sample from each state, based on weights
state_weights <- state_weights %>%
  mutate(sample_size = round(weight * N))

set.seed(10049)
overall_samples <- NULL
for (i in 1:nrow(state_weights)) {
  state <- state_weights[i, ]$state
  size <- state_weights$sample_size[i]
  sample <- overall_dat %>% filter(state == state) %>% 
    sample_n(size, replace = FALSE)
  overall_samples <- rbind(overall_samples, sample)
}

overall_samples <- overall_samples %>% group_by(state) %>% 
  mutate(schoolID = row_number()) %>% 
  relocate(schoolID, .after = stateID) %>% 
  ungroup() %>% 
  arrange(stateID)
```



# EDA

## Data Visualization

### Vaccination Rates by States 

```{r}
rate_long <- clean_data %>% 
  select(state_abbr, mmr, overall) %>%
  pivot_longer(cols = c(mmr, overall), names_to = "vaccine_type", values_to = "rate")

rate_long %>% 
  filter(state_abbr %in% no_mmr) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "Overall Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(state_abbr %in% no_overall) %>% 
  ggplot(aes(y= rate, x = state_abbr)) +
  geom_boxplot() +
  labs(x = "State", y = "MMR Rate") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))

rate_long %>% 
  filter(!(state_abbr %in% c(no_mmr, no_overall))) %>% 
  ggplot(aes(y= rate, x = state_abbr, color = vaccine_type)) +
  geom_boxplot() +
  labs(x = "State", y = "rate_long") +
  scale_color_manual(values = c("overall" = "blue", "mmr" = "red")) + 
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))
```


### State School Aged Population and Per Capita

```{r}
clean_data %>% 
  ggplot(aes(x = state_abbr, y = schagepop2016)) +
  geom_point(size = 3, color = "red") + 
  geom_segment(aes(x = state_abbr, xend = state_abbr, y = 0, yend = schagepop2016)) +
  labs(x = "State", "School-Aged Population")+
  theme_minimal() 

clean_data %>% 
  ggplot(aes(x = state_abbr, y = per_capita)) +
  geom_point(size = 3, color = "red") + 
  geom_segment(aes(x = state_abbr, xend = state_abbr, y = 0, yend = per_capita)) +
  labs(x = "State", y="Per Capita")
  theme()
```


### Vaccination rate vs. School, County, City, Capita, and Spending 

```{r}
state_summary <- clean_data %>% group_by(state) %>% 
  summarise(Total_schools = n(), 
            n_county = ifelse(sum(is.na(county)) == Total_schools, 0, n_distinct(county)), 
            n_county_na = sum(is.na(county)),
            n_city = ifelse(sum(is.na(city)) == Total_schools, 0, n_distinct(city)), 
            n_city_na = sum(is.na(city)), 
            avg_mmr = mean(mmr, na.rm = TRUE), 
            avg_overall = mean(overall, na.rm = TRUE), 
            avg_enroll = mean(enroll)) %>% 
  arrange(avg_enroll) %>% 
  left_join(clean_data %>% group_by(state) %>% 
          select(state, year, per_capita, schagepop2016, statespending2016), by="state")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = Total_schools, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Schools in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_county, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Counties in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_mmr, color =year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = n_city, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Total number of Cities in a State", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = per_capita, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "Per Capita", 
       y = "Average Overall Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_mmr, color = year))+
  theme_minimal()+
  labs(x = "State Spending in 2016", 
       y = "Average MMR Vaccination Rate")

state_summary %>% 
  ggplot()+
  geom_point(aes(x = statespending2016, y = avg_overall, color = year))+
  theme_minimal()+
  labs(x = "State Spending 2016", 
       y = "Average Overall Vaccination Rate")
```

### Florida 

```{r}
clean_data %>% filter(state == "Florida") %>% select(overall) %>% summary()
```


## EDA for MMR Data 

```{r}
mmr_df %>% 
  group_by(state) %>% 
  summarise(n_school = n_distinct(name),
            min_mmr = min(mmr, na.rm=TRUE),
            mean_mmr = mean(mmr, na.rm=TRUE), 
            median_mmr = median(mmr, na.rm=TRUE), 
            max_mmr = max(mmr, na.rm=TRUE), 
            mmr_na = sum(is.na(mmr)), 
            year = unique(year)) %>% 
  arrange(year) %>% 
  kable(booktabs = TRUE, caption = "The summary of MMR vaccination rates", 
        col.names = c("State", "No.School", "Min", "Mean", "Median", "Max", "No.Missing", "School Year")) %>% 
  kable_styling(latex_options = c("striped", "scale_down")) %>% 
  landscape()
```



## EDA for Overall Data 


# Statistical Analysis 

The data has many schools within each state, meaning that each state has repeated measurements by schools. Therefore, the vaccination rate between schools within the same state are not independent of each other, violating the assumption of independence of standard regression models. Therefore, we will model the covariance structures with Generalized Least Squares. 

## MMR 

Let $Y_{ij}$ denote the MMR vaccination rate of the $i^{th}$ state and $j^{th}$ school. 

$$
E(Y_{ij}) = \beta_0 + \beta_1 \text{Capita}_i + \beta_2\text{NSchool}_i+\beta_3 \text{Year}_i
$$
where Capita$_i$ was calculated by dividing the state spending by school aged population, `NSchool` represents the number of schools in each state, and `Year` denotes the school-year that the observation is recorded (1 = 2018-19, 0 = 2017-18). 




#### Model comparison

```{r, eval=FALSE}
outAIC <- function(x){
  sumx <- summary(x)
  out <- c(sumx$AIC, sumx$BIC)
  names(out) <- c("AIC", "BIC")
  out
}

allaic <- rbind(
  Unstructured = outAIC(mmr_unstr), 
  CompoundSymmetry = outAIC(mmr_compsym), 
  AR1 = outAIC(mmr_ar1), 
  Exponential = outAIC(mmr_expo)
)

allaic
```




### GEE 

```{r}
library(geepack)

mmr_samples$year <- as.factor(mmr_samples$year)

# Fit the GEE model
gee_model.ind1 <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "independence")

# Summary of the model
summary(gee_model.ind1)$coef

# Fit the GEE model
gee_model.ex1 <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "exchangeable")

# Summary of the model
summary(gee_model.ex1)$coef


# Fit the GEE model
gee_model.ar1 <- geeglm(mmr ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = mmr_samples, 
                    id = stateID, 
                    corstr = "ar1")

# Summary of the model
summary(gee_model.ar1)$coef
```



```{r}
summary(gee_model.ar1)$corr
```

```{r}
summary(gee_model.ar1)$corr
summary(gee_model.ex1)$corr
```

```{r}
QIC(gee_model.ar1,gee_model.ex1,gee_model.ind1) %>%  # default
  pander::pander(caption = "QIC")
```




```{r}
library(geepack)

overall_samples$year <- as.factor(overall_samples$year)

# Fit the GEE model
gee_model.ind2 <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "independence")

summary(gee_model.ind2)$coef

# Fit the GEE model
gee_model.ex2 <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "exchangeable")

# Summary of the model
summary(gee_model.ex2)$coef

# Fit the GEE model
gee_model.ar12 <- geeglm(overall ~ per_capita + n_school + type, 
                    family = gaussian, 
                    data = overall_samples, 
                    id = stateID, 
                    corstr = "ar1")

# Summary of the model
summary(gee_model.ar12)$coef
```

```{r}
QIC(gee_model.ar12,gee_model.ex2,gee_model.ind2) %>%  # default
  pander::pander(caption = "QIC")
```
```{r,echo=FALSE}
library(knitr)
library(kableExtra)
library(rmarkdown)

results_df <- data.frame(
  Term = c("Intercept", "per_capita", "n_school", "typeOthers", "typePrivate", "typePublic", "typeUnknown"),
  MMR_Independence = c("85.597 (p < 0.001)", "0.0004 (p = 0.00021)", "0.00059 (p = 0.01595)", 
                       "4.678 (p < 0.001)", "-2.687 (p < 0.138)", "2.672 (p < 0.102)", 
                       "2.7628 (p < 0.0334)"),
  MMR_Exchangeable = c("8.48e+01 (p < 0.001)", "3.76e-04 (p = 0.00805)", 
                       "3.77e-04 (p = 0.17570)", "6.13e+00 (p < 0.001)", 
                       "0.432 (p = 0.71506)", "5.31e+00 (p < 0.001)", 
                       "3.32e+00 (p = 0.01345)"),
  MMR_AR1 = c("84.9307 (p < 0.001)", "0.000394 (p = 0.00229)", "0.000587 (p = 0.016673)",
              "5.40787 (p < 0.001)", "-1.699691 (p = 0.278339)", "3.459308 (p = 0.009406)", 
              "3.546702 (p = 0.003588)"),
  
  Overall_Independence = c("77.623 (p < 0.001)", "0.0011 (p = 0.00293)",
                           "0.00105 (p < 0.001)", '-',"-6.151 (p < 0.001)", 
                           "-1.989 (p = 0.00209)", "-4.032 (p = 0177)"),
  Overall_Exchangeable = c("77.77895 (p < 0.001)", "0.00109 (p = 0.00326)", 
                           "0.00103 (p < 0.001)", '-',"-6.06678 (p < 0.001)", 
                           "-1.91152 (p = 0.0305)", "-3.86783 (p = 0.201)"),
  Overall_AR1 = c("77.62593 (p < 0.001)", "0.00110 (p = 0.00294)", 
                  "0.00105 (p < 0.001)", "-", "-6.15253 (p < 0.001)", 
                  "-1.98606 (p = 0.02102)", "-4.02961 (p = 0.178)"),
  stringsAsFactors = FALSE
)
# Create the table with kable and kableExtra
kable(results_df, booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(font_size = 4.7, latex_options = "striped", position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c("Parameter Estimates for Model with Outcome MMR" = 4, 
                     "Parameter Estimates for Model with Outcome Overall"=3))
```
```{r}
vac=read.csv('vaccine.csv')
vac_new = vac  %>% 
  group_by(state) %>% 
  mutate(v=mean(vac$overall[!is.na(vac$overall)])) 
l=lm(v~statespending2016+state,vac_new)
summary(l)
plot(l)
```

